Covariate analysis of MAYO reprocessed transcripts (with TMM normalisation and covariates adjusted)

### 2017-09-08 code/Rmd/MAYO_REPROCESSED_ISOFORM.Rmd
Added markdown to process MAYO isoform

---
title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)
library(glmnet)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn8672415'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';

thisFileName <- 'MAYO_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn8690799';#'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn8690904';#'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

@@

# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']

@@

# Get picard metrics from synapse
METADATA.PICARD_CE_ID <- 'syn8698214';#'syn8380392'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
# METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
# METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
# 
# METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
#   gsub('/picard.CombinedMetrics.csv','',.) %>%
#   sapply(function(x){str_split(x, '\\.')[[1]][1]})
METADATA.PICARD_CE <- CovariateAnalysis::downloadFile(METADATA.PICARD_CE_ID) %>%
  dplyr::rename(ID = sample)
colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))

METADATA.PICARD_TC_ID <- 'syn8698211';#'syn8383233'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
# METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
# METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
# METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
#   gsub('/picard.CombinedMetrics.csv','',.) %>%
#   sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
#   gsub('Aligned','',.)

METADATA.PICARD_TC <- CovariateAnalysis::downloadFile(METADATA.PICARD_TC_ID) %>%
  dplyr::rename(ID = sample)
colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))

# Merge all metadata
METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  inner_join(rbindlist(list(METADATA.PICARD_TC, METADATA.PICARD_CE), use.names = T, fill = T)) %>%
  as.data.frame

# Get gene specific parameters from synapse
GENE.PARAM = downloadFile('syn8449369')
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')

GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id

@@

### Predict missing PMI
```{r predict.pmi}
# Get PMI
PMI = METADATA[,c('PMI', 'ID', 'Donor_ID'), drop = F]
PMI$PMI = PMI$PMI

# Get expression (convert counts to cpm)
expr = voom(COUNT, design = NULL)$E
ind = which(rowSums(expr>=1)/dim(expr)[2] >= 0.5)
expr = expr[ind,]

# Combine expression from different regions of brain of an individual into one
expr = plyr::dlply(PMI, .(Donor_ID), .fun = function(x, expr){
  data.frame(value = rowMeans(expr[,x$ID,drop = F], na.rm = T)) %>%
    plyr::rename(c('value' = unique(x$Donor_ID)))
}, expr, .parallel = F) %>%
  do.call(cbind,.)

# Split data into trainining and prediction set
PMI = dplyr::select(PMI, Donor_ID, PMI) %>% unique()
rownames(PMI) = PMI$Donor_ID
PMI = PMI[colnames(expr),]

# Fit a logistic regression model
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]
y.train = y.train[colnames(x.train), 'PMI']

plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
R2.allgenes = 1-(ssres/sstot)

y.predict.all = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('Donor_ID') %>%
  plyr::rename(c('1' = 'PMI'))

# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F) %>%
  dplyr::mutate(ID = x) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')

# Initial normalisation of gene expression
ind = rownames(expr) %in% CONTROL.PROBES$ensembl_gene_id 
expr = expr[ind,]

# Fit a logistic regression model
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]
y.train = y.train[colnames(x.train), 'PMI']

plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
R2.smallgenes = 1-(ssres/sstot)

y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('Donor_ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = PMI %>% 
  dplyr::select(Donor_ID, PMI) %>%
  dplyr::filter(!is.na(PMI)) %>%
  list(y.predict) %>%
  rbindlist(use.names = T, fill = T)
PMI$PMI[PMI$PMI < 0] = 0

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```
Fitted R2 values are, (i) with all gx features `r round(R2.allgenes, 3)`, (ii) with small set of inferred gx features `r round(R2.smallgenes,3)`

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("BrainRegion.Diagnosis", "BrainRegion.ApoE")
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

```{r data.explore, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme

# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme

# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme

multiplot(plotlist = p, cols = 3)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)

# Check gene biotype
## Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

## Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
                       filters = "ensembl_gene_id", 
                       values = PROCESSED_COUNTS$filteredExprMatrix$genes,
                       mart = mart)

summary(factor(Ensemble2HGNC$gene_biotype)) %>%
  rownameToFirstColumn('Biotype') %>%
  dplyr::rename(fraction = DF) %>%
  dplyr::mutate(fraction = fraction/dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]) %>%
  dplyr::filter(fraction >= 0.01) %>%
  kable
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

@@

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

@@

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
CQN.GENE_EXPRESSION$E = LOG.CPM
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

@@

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID,
                                             method = 'lmer')
  # correlation = list(); correlation$cor = 0.457197

@@

    # Fit linear model using new weights and new design
    VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                              design = DM1$design,
                              weights = VOOM.GENE_EXPRESSION$weights,
                              block = COVARIATES$Donor_ID, 
                              correlation = correlation$cor)

    # Residuals after normalisation
    RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                  CQN.GENE_EXPRESSION$E)
  } else {
    notEstimable = c(notEstimable, postAdjustCovars[length(postAdjustCovars)])
    postAdjustCovars = postAdjustCovars[1:(length(postAdjustCovars)-1)]
  }

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
                                           block = COVARIATES$Donor_ID,
                                           method = 'lmer')

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design,
            block = COVARIATES$Donor_ID, 
            correlation = correlation$cor)

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref) %>%
  group_by(BrainRegion.ref, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
                                           block = COVARIATES$Donor_ID,
                                           method = 'lmer')

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design,
            block = COVARIATES$Donor_ID, 
            correlation = correlation$cor)

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
  dplyr::mutate(reference = paste0('ApoE',reference),
                against = paste0('ApoE',against)) %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

>>>>

---
title: "Covariate analysis of MAYO reprocessed transcripts (with TMM normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED_ISOFORM.Rmd",
                                 parentId = "syn8672415",
                                 entityName = "MAYO Reprocessed Isoform (Normalised)")
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn8672415'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed transcript data';

thisFileName <- 'MAYO_REPROCESSED_ISOFORM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_isoform')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn10507725';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn10499229';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

@@

@@

# Get PMI estimates from gene expression analysis
METADATA_PMI = downloadFile('syn8466814') %>%
  dplyr::select(SampleID, PMI) %>%
  dplyr::rename(ID = SampleID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = 'syn8466814'

# Merge all metadata
METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  left_join(METADATA_PMI) %>%
  as.data.frame

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("BrainRegion.Diagnosis", "BrainRegion.ApoE")
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

@@

```{r data.explore, fig.width = 12, fig.height = 5}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

multiplot(plotlist = p, cols = 3)
```
### Filter transcripts
Remove genes that have less than 5 cpm counts in at least 50% of samples per BrainRegion x Diagnosis
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=5, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

@@

### Library Normalisation
Initial library normalisation usign tmm
```{r tmm}
# Compute offset for gene length and gc content
TMM.GENE_EXPRESSION = calcNormFactors(NEW.COUNTS, method = 'TMM')
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = NULL, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

@@

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
VOOM.GENE_EXPRESSION$E = LOG.CPM
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = VOOM.GENE_EXPRESSION$E
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID,
                                             method = 'lmer')
  # correlation = list(); correlation$cor = 0.457197

@@

    # Fit linear model using new weights and new design
    VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION$E,
                              design = DM1$design,
                              weights = VOOM.GENE_EXPRESSION$weights,
                              block = COVARIATES$Donor_ID, 
                              correlation = correlation$cor)

    # Residuals after normalisation
    RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                  VOOM.GENE_EXPRESSION$E)
  } else {
    notEstimable = c(notEstimable, postAdjustCovars[length(postAdjustCovars)])
    postAdjustCovars = postAdjustCovars[1:(length(postAdjustCovars)-1)]
  }

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
                                           block = COVARIATES$Donor_ID,
                                           method = 'lmer')

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor,
                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design,
            block = COVARIATES$Donor_ID, 
            correlation = correlation$cor)

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('Tx.ID')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-')
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(Tx.ID, Comparison, Direction, BrainRegion.ref) %>%
  group_by(BrainRegion.ref, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
                                           block = COVARIATES$Donor_ID,
                                           method = 'lmer')

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor,
                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design,
            block = COVARIATES$Donor_ID, 
            correlation = correlation$cor)

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('Tx.ID')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
  dplyr::mutate(reference = paste0('ApoE',reference),
                against = paste0('ApoE',against)) %>%
  tidyr::unite(Comparison, reference, against, sep = '-')
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
